A formal treatment of bidirectional typing

### 2024-01-11 tex/ESOP24/S3-defs.tex
[ tex ] Some revisions before removing appendix for the camera ready version

\begin{example}
  Observe that $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
  \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
  \renewcommand{\eqnannotationfont}{\rmfamily\footnotesize} % package default
  \vspace{2em}
  \begin{equation*}
    \frac{\Gamma, \eqnmarkbox[blue]{ext1}{\isTerm{x} : A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
    \qquad\text{and}\qquad
    \frac{\Gamma, \eqnmarkbox[blue]{ext2}{\cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty2}{A \bto B} \qquad \Gamma, \eqnmarkbox[blue]{ext3}{\cdot} \vdash \isTerm{u} : \eqnmarkbox[red]{aty3}{A}}{\Gamma \vdash \isTerm{t\;u} : \eqnmarkbox{ty2}{B}}
  \end{equation*}
  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension context}
  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument type}
  \annotate[yshift=0em]{below}{ty1,ty2}{target type}
  \vspace{.5em}
  \\
  if the empty extension context $\cdot$ is added verbosely, so they can be specified by binding arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$ respectively.
\end{example}

@@

\begin{figure}
  \centering
  \small
  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for a language $(\Sigma, \Omega)$ with free variables in~$V$}
  \begin{mathpar}
    \inferrule{\isTerm{x} \in V}{V |-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotatecolon A}}\,\Rule{Anno}
    \\
    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n} } {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op} \and
    \text{for $\bop$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms}
  \label{fig:raw-terms}
\end{figure}

@@

Before defining typing derivations, we need a definition of typing contexts.

>>>>

\begin{example}
  Observe that $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
  \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
  \renewcommand{\eqnannotationfont}{\rmfamily\footnotesize} % package default
  \vspace{2em}
  \begin{equation*}
    \frac{\Gamma, \isTerm{x} : \eqnmarkbox[blue]{ext1}{A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
    \qquad\text{and}\qquad
    \frac{\Gamma, \eqnmarkbox[blue]{ext2}{\cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty2}{A \bto B} \qquad \Gamma, \eqnmarkbox[blue]{ext3}{\cdot} \vdash \isTerm{u} : \eqnmarkbox[red]{aty3}{A}}{\Gamma \vdash \isTerm{t\;u} : \eqnmarkbox{ty2}{B}}
  \end{equation*}
  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension context}
  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument type}
  \annotate[yshift=0em]{below}{ty1,ty2}{target type}
  \vspace{.5em}
  \\
  if the empty context $\cdot$ is added verbosely, so they can be specified by arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$, respectively, with $\Type_{\Sigma_{\bto}}(A, B)$ as types.
\end{example}

@@

@@

\begin{figure}
  \centering
  \small
  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for a language $(\Sigma, \Omega)$ with free variables in~$V$}
  \begin{mathpar}
    \inferrule{\isTerm{x} \in V}{V |-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotatecolon A}}\,\Rule{Anno}
    \\
    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n} } {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op} \and
    \text{for $\bop$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms}
  \label{fig:raw-terms}
\end{figure}
Before defining typing derivations, we need a definition of typing contexts.

### 2024-01-11 tex/ESOP24/S5-synthesis.tex
[ tex ] Rewrite the example for binding arity with colourful annotations
[ tex ] Some revisions before removing appendix for the camera ready version

\begin{defn}\label{def:mode-correctness-args}
  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a predicate defined by
  {\small
  \begin{align*}
    \MCas( \cdot ) & = \top \\
    \MCas\left(\biargvec, \chkbiarg[n]\right)
                                  & = \fv(\Delta_n, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
                                  & = \phantom{, A_n} \fv(\Delta_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
  \end{align*}}
  where $\MCas( \cdot ) = \top$ means that an empty list is always mode-correct.
\end{defn}
This definition encapsulates the idea that every `input' type variable, possibly derived from an extension context~$\Delta_n$ or a checking argument~$A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must also be met for every tail of the argument list to ensure that `output' variables accessible at each argument are from preceding arguments only, hence an inductive definition.
\begin{defn}\label{def:mode-correctness}
  An operation $\biop$ is \emph{mode-correct} if 
  \begin{enumerate}
    \item either $d$ is $\chk$, its argument list is mode-correct with respect to $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
    \item or $d$ is $\syn$, its argument list is mode-correct with respect to $\emptyset$, and $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{defn}
For a checking operation, an `input' variable of an argument could be derived from~$A_0$, as these are known during type checking as an input.
Since every inhabitant of~$\Xi$ can be located in either~$A_0$ or synthesised variables, we can determine a concrete type for each inhabitant of~$\Xi$ during type synthesis.
On the other hand, for a synthesising operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to~$\emptyset$.
Also, the set of synthesised variables alone should include every type variable in~$\Xi$ and particularly in~$A_n$.

@@

It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for the simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
\begin{lemma}\label{lem:decidability-mode-correctness}
  For any operation $\biop$, it is decidable whether $o$ is mode-correct.
\end{lemma}

Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction on the typing derivations.
However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
\begin{lemma}[Uniqueness of synthesised types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true.
        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
      \end{enumerate}
  \end{itemize}\qed
\end{proof}

>>>>

\begin{defn}\label{def:mode-correctness-args}
  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a predicate defined by
  {\small
  \begin{align*}
    \MCas( \cdot ) & = \top \\
    \MCas\left(\biargvec, \chkbiarg[n]\right)
                                  & = \fv(\Delta_n, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
                                  & = \phantom{, A_n} \fv(\Delta_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
  \end{align*}}
  where $\MCas( \cdot ) = \top$ means that an empty list is always mode-correct.
\end{defn}
This definition encapsulates the idea that every `input' type variable, possibly derived from an extension context~$\Delta_n$ or a checking argument~$A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must also be met for every tail of the argument list to ensure that `output' variables accessible at each argument are from preceding arguments only, hence an inductive definition.
\begin{defn}\label{def:mode-correctness}
  An arity $\biarity$ is \emph{mode-correct} if 
  \begin{enumerate}
    \item either $d$ is $\chk$, its argument list is mode-correct with respect to $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
    \item or $d$ is $\syn$, its argument list is mode-correct with respect to $\emptyset$, and $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if every operation's arity is mode-correct.
\end{defn}
For a checking operation, an `input' variable of an argument could be derived from~$A_0$, as these are known during type checking as an input.
Since every inhabitant of~$\Xi$ can be located in either~$A_0$ or synthesised variables, we can determine a concrete type for each inhabitant of~$\Xi$ during type synthesis.
On the other hand, for a synthesising operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to~$\emptyset$.
Also, the set of synthesised variables alone should include every type variable in~$\Xi$ and particularly in~$A_n$.

@@

It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for the simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
\begin{lemma}\label{lem:decidability-mode-correctness}
  For any bidirectional binding arity $\biarity$, it is decidable whether it is mode-correct.
\end{lemma}

Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction on the typing derivations.
However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
\begin{lemma}[Uniqueness of synthesised types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotatecolon A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true.
        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
      \end{enumerate}
  \end{itemize}\qed
\end{proof}

### 2024-01-11 tex/ESOP24/S8-future.tex
[ tex ] Some revisions before removing appendix for the camera ready version

\section{Discussion} \label{sec:future}
We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and aspects remain untouched.
In this section we discuss related work and potential directions for future research.
\subsubsection{Language formalisation frameworks}

The study of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which is used to prove a general confluence theorem.
Many general definitions and frameworks for defining logics and type theories have been proposed, and we classify them into two groups by where signatures reside: the meta level or the object level of a meta-language:
\begin{enumerate}
  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
\end{enumerate}

@@

\subsubsection{Beyond syntax-directed systems}

For non-syntax-directed systems, since there can be multiple moding or typing rules for a language construct, both the mode decorator and the bidirectional type synthesiser will need to backtrack, and the decorator may need to produce all possible mode derivations so as not to limit the synthesiser's choices.
For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and typing derivations, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer be in perfect correspondence.
Completeness may need to be reformulated to take all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation.

@@

\begin{credits}
\subsubsection{Acknowledgements.}
We thank Kuen-Bang {Hou (Favonia)} and anonymous reviewers for their comments and suggestions.
\end{credits}

>>>>

\section{Discussion} \label{sec:future}
We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and aspects remain untouched.
%In this section we discuss related work and potential directions for future research.
\subsubsection{Language formalisation frameworks}

The study of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
Many general definitions and frameworks for defining logics and type theories have been proposed, and we classify them into two groups by where signatures reside: the meta level or the object level of a meta-language:
\begin{enumerate}
  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
\end{enumerate}

@@

\subsubsection{Beyond syntax-directedness}

To explore general settings, an extreme possibility is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.

We could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is allowed to have multiple bidirectional variants, including different orders of their premises.
In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
Completeness could still take the simple form presented in this paper too.

The next step is to consider non-syntax-directed systems where each typing rule has multiple bidirectional variants.
For this setting, the mode decorator and the bidirectional type synthesiser will have to backtrack.
It is still possible to treat soundness as the separation of mode and type information, but completeness will pose a problem.
For every raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
To retain the mode decoration, one could make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of mode derivations into a bidirectional typing derivation.
%It should be possible to fix this problem while retaining the mode decoration phase.

@@

\begin{credits}
\subsubsection{Acknowledgements.}
We thank Kuen-Bang {Hou (Favonia)} and anonymous reviewers for their comments and suggestions.
The work was supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.
\end{credits}
